\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-26 chapters/abstract_fr.tex
All review comments
Layout chapter gaze independence

\chapter*{R\'esum\'e}
Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a palier cette limitation par améliorer des méthodes de BCI
indépendantes du regard et la décodage de la attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation pour la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

@@

\paragraph{Mots-clés :}
interfaces neuronales directes,
\'electroenc\'ephalographie,
potentiels \'evoqu\'es,
attention visuo-spatiale cachée,
decodage (multi-)linéaire,
handicap\'es moteurs

>>>>

\chapter*{R\'esum\'e}
\addcontentsline{toc}{chapter}{French abstract}

Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a palier cette limitation par améliorer des méthodes de BCI
indépendantes du regard et la décodage de la attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation pour la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

@@

\paragraph{Mots-clés :}
interfaces neuronales directes,
\'electroenc\'ephalographie,
potentiels \'evoqu\'es,
attention visuo-spatiale,
decodage (multi-)linéaire,
handicap\'es moteurs

### 2024-11-26 chapters/gaze_independence.tex
Layout chapter gaze independence

\chapter{Gaze-independent visual \acsp{bci}}
\label{sec:gaze-independent}
\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

\noindent\emph{The last paragraph of \cref{sec:gaze-independence/sota},
\cref{sec:gaze-independence/oculomotor/gaze}, and the first paragraph of
\cref{sec:gaze-independence/objectives} were adapted from
\textcite{VanDenKerchove2024}.}

\section{Eye motor impairment in individuals with \acl{sspi}}%
\label{sec:gaze-dependence}

@@

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010} stimulation do not rely on gaze redirection
but also result in lower \acp{itr} compared to visual paradigms.
A survey in by \textcite{Riccio} from 2012 reported that \acp{itr} ranged between
1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.

On the tactile front, recent studies use steady-state somatosensory evoked
potentials~\cite{Petit2021} or tactile oddball paradigms~\cite{Waal2012}
These are analogous to the \ac{ssvep} and visuall oddball paradigms,
but with vibrotactile stimulation at different spatial locations and/or
frequencies~\cite{Han2020}.
Some noteworthy advancements achieved an \ac{itr} of 14.77
bits/min~\cite{Liu2019} and 20.73 bits/min~\cite{Herweg2016} with tactile
oddball stimulation through highly optimized stimulation?
Still, other recent tactile paradigm studies~\cite{Herweg2016a,Han2020,Jin2020,Li2019,Eidel2022}
fail to exceed 5 bits/min.

@@

While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that it can provide added value in gaze-independent
settings when coupled with visual oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Auditory and tactile stimulation can even be paired to create a fully
non-visual, hybrid \ac{bci}, with a reported \ac{itr} of 11.66 bits/min~\cite{Zhang2020}.

Yet, auditory and somatosensory \acp{bci} also suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.
However, non-visual stimulation modalities are still valuable for developing
BCIs for individuals affected with severe vision loss or
blindness~\cite{Nijboer2008,Rutkowski2015}, which can occur in some cases of
\ac{lis}~\cite{Patterson1986}.

@@

This leads us to the second approach: current visual stimulation paradigms can be optimized so that
the stimuli are always present in the field of view, either overtly~\cite{Acqualagna2013, Won2018, Lin2018} or covertly~\cite{Pires2011,Lees2018}.
Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm, in particular, is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the user's field of view.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small
sets of targets (e.g. 3) are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

@@

This leads us to adopt the following hypothesis: \emph{it is possible to improve
the (gaze-independent) accuracy of a spatially organized visual oddball
\ac{erp}-based brain computer interface by using a suited decoding strategy}.

@@

A widely used method is \ac{hoda}~\cite{Phan2010}, which projects tensors onto subspaces
optimized for class separation, similar to \ac{lda} but extended to multidimensional data.
This method has been applied to various \ac{bci} tasks, including \ac{erp} and
\ac{mi} decoding, enabling the extraction of class-relevant features while reducing
the data's dimensionality.

@@

\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
    \caption[Overt \ac{vsa}]{%
      \emph{Overt} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
    \caption[Covert \ac{vsa}]{%
      \emph{Covert} \ac{vsa}.
      Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
    }
    \label{fig:gaze/vsa/covert}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      The user is free to direct their gaze as they deem most comfortable.
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Acl{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze/vsa}
\end{figure}

>>>>

\chapter{Gaze-independent visual \acsp{bci}}
\label{sec:gaze-independent}
\epigraph{%
  \justifying
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

\publicationnote{The last paragraph of \cref{sec:gaze-independence/sota},
\cref{sec:gaze-independence/oculomotor/gaze}, and the first paragraph of
\cref{sec:gaze-independence/objectives} were adapted from
\textcite{VanDenKerchove2024}.}

\section{Eye motor impairment in \acs{bci} users}%
\label{sec:gaze-dependence}

@@

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010} stimulation do not rely on gaze redirection
but also result in lower \acp{itr} compared to visual paradigms.
A survey by \textcite{Riccio2012} from 2012 reported that \acp{itr} ranged between
1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.

On the tactile front, recent studies use steady-state somatosensory evoked
potentials~\cite{Petit2021} or tactile oddball paradigms~\cite{Waal2012}
These are analogous to the \ac{ssvep} and visuall oddball paradigms.
Instead of visual stimulation, however, they use physical vibrotactile pulses
stimulation at different spatial locations and/or frequencies~\cite{Han2020}.
Some noteworthy advancements have achieved an \ac{itr} of 14.77 bits/min~\cite{Liu2019}
and 20.73 bits/min~\cite{Herweg2016} with a tactile oddball paradigm through
highly optimized stimulation.
Still, other recent tactile \ac{bci} studies~\cite{Herweg2016a,Han2020,Jin2020,Li2019,Eidel2022}
fail to exceed 5 bits/min.

@@

While auditory and somatosensory stimulation on their own might yield poor
\ac{itr}, \textcite{Yin2016} showed that these modalities can provide added
value in gaze-independent settings when coupled with visual oddball
stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94 bits/min.
In general, their approach is interesting as it follows a philosophy that
relies on establishing as many communication pathways as possible.
Auditory and tactile stimulation can even be paired to create a fully
non-visual, hybrid \ac{bci}, with a reported \ac{itr} of 11.66 bits/min~\cite{Zhang2020}.

Yet, auditory and somatosensory \acp{bci} also suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.
However, non-visual stimulation modalities are still valuable for developing
\acp{bci} for individuals affected with severe vision loss or
blindness~\cite{Nijboer2008,Rutkowski2015}, which can occur in some cases of
\ac{lis}~\cite{Patterson1986}.

@@

This leads us to the second approach: current visual stimulation paradigms can be optimized so that
the stimuli are always present in the field of view, either overtly~\cite{Acqualagna2013, Won2018, Lin2018} or covertly~\cite{Pires2011,Lees2018}.
Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm, in particular, is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the user's field of view.

We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small
sets of targets (e.g. 3) are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

@@

This leads us to adopt the following central hypothesis:
\emph{it is possible to improve the (gaze-independent) accuracy of a spatially
organized visual oddball \ac{erp}-based brain computer interface by using a
suited decoding strategy}.

@@

A relatively well-known example is \ac{hoda}~\cite{Phan2010}
It projects tensors onto subspaces optimized for class separation, similar to \ac{lda} but extended to multidimensional data.
\Ac{hoda} has been applied to various \ac{bci} tasks, including \ac{erp} and
\ac{mi} decoding, enabling the extraction of class-relevant features while reducing
the data's dimensionality.

@@

\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.pdf}
    \caption[Overt \ac{vsa}]{%
      \emph{Overt} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.pdf}
    \caption[Covert \ac{vsa}]{%
      \emph{Covert} \ac{vsa}.
      Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
    }
    \label{fig:gaze/vsa/covert}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.pdf}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.pdf}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      The user is free to direct their gaze as they deem most comfortable.
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Acf{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm
    interface. Full lines indicated gaze fixation, dashed lines \ac{vsa}.
  }
  \label{fig:gaze/vsa}
\end{figure}
